---
title: "Planning weekly food"
subtitle: "using R to menu plan and create a food budget"
author: "Gerhard Groenewald"
date: '2018-12-15'
output:
  html_document: default
  pdf_document: default
layout: post
# draft: true
tags:
- tips tricks
- intermediate
- R
- development
- data science
- analysis
- management

comments: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
A few years ago I woke up to an epiphany, realising that I was becoming my dad.
I had started a campaign of dealing with **wastefulness**, switching off lights and eating leftovers to name but a few examples.  I set out to transform our *menu planning* and *the weekly food shop* as part of this crusade.

Menu planning is a chore which comes easily to some. For others like me, though, it is just another thing to think about on top of an already busy life.  Admittedly, it gets easier once establishing a routine, but holidays and the like upsets the momentum and things move back to square one. 

Food planning, especially for a young family is essential, without which it can get out of hand and quickly become expensive and wasteful.  

I decided that it was not going to beat me!  I reached into my bag of tricks and conjured a helper to tackle this quest. A simple piece of R code creates a randomised 'inspiration' menu plan.  It meant that I had to document a few dozen recipes to start, but the effort soon paid off.

It is now a tonne easier to plan our weekly menu plan, and we tend not to use the program anymore.  However, it served as a bootstrap, transforming food planning into what is now a routine and straightforward task.<br><br>

This post covers and explains the code, and how to use its output to tack on budgeting for example.<br><br><hr><br>

## Code

### Library
Load the packages used in the demonstration.<br><br>

```{r}
package_list <- c("tidyverse", "readxl", "lubridate", "knitr", "kableExtra")
invisible(suppressPackageStartupMessages(lapply(package_list, library, character.only = TRUE)))
rm(package_list)

```

### Functions
The functions below primarily helps with presentation and formatting.  The inline comments below describe the working and sequence of each function.  I have [previously written](/post/2018-11-08-summarising_tables/#functions) about these functions and how they are used.<br><br>

```{r}
func_rename <-
  function(x) {
    
    # sequence is important here
    # regex removing all chars not alpha-numeric, underscore or period
    gsub("[^[:alnum:] \\_\\.]", "", x) %>%
      
      # replace any space or period with underscore
      str_replace_all(pattern = " |\\.", replacement = "\\_") %>%
      
      # replace multiple underscores with one
      str_replace_all(pattern = "\\_+", replacement = "\\_") %>%
      
      # remove trailing underscores and period
      str_remove(pattern = "\\_$|\\.$")
  }

func_legible_boolean <-
  
  # returns `feature name` when TRUE and "Not_" + `feature name` otherwise
  function(df) {
    
    # add row numbers to input dataframe
    df <-
      df %>%
      ungroup() %>%
      mutate(row_id = row_number())
    
    # select row_id and other logical fields
    tmp <-
      df %>%
      select(row_id, select_if(., is.logical) %>% names(.)) %>%
      
      # gather all logical fields and keep row_id to retain identity
      gather(key, value,-row_id) %>%
      
      # when value is TRUE then return the feature name else "Not_" pasted in
      # front of it
      mutate(value = case_when(
        value ~ key,
        TRUE ~ paste("Not", key, sep = "_")
      )) %>%
      
      # return the feature names back to header positions
      spread(key, value)
    
    vars <- 
      df %>%
      select_if(is.logical) %>%
      names()
    
    df %>% 
      
      # return all values from the table after excluding the original logical
      # fields and join the newly adjuted features back into the table using the
      # row_id
      select(-(!!vars)) %>%
      inner_join(tmp,
                 by = "row_id") %>%
      
      # remove the row_id as the join is complete
      select(-row_id)
  }

func_create_summary <-
  
  # group by factors, summarising numeric fields to sum total
  function(df) {
    
    # create a list of factors from the input dataframe
    tmp_factors <- df %>% select_if(is.factor) %>% names()
    
    # mutate all factors in the dataframe to characters
    df <- df %>% ungroup() %>% mutate_if(is.factor, as.character)
    
    
    # list original dataframe with a newly created summary row
    list(
      
      # original input dataframe
      df,
      
      # summary row, replaceing each factor value to "Total"
      df %>%
        mutate_at(tmp_factors, ~"Total") %>%
        
        # group by for each of the original factor features
        group_by_at(tmp_factors) %>%
        
        # summarise all numeric values to sum total, ignoring NULL values
        summarise_if(is.numeric, sum, na.rm = TRUE)
    ) %>% 
      
      # union all tables contained in the list, including original dataframe and
      # newly created summary row
      reduce(union_all) %>%
      
      # change factors back into factors
      mutate_at(tmp_factors, factor) %>% 
      return()
  }

func_present_headers <-
  function(.) {
    
    # replace all `spaceholder underscores` with spaces
    str_replace_all(., pattern = "_", replacement = " ") %>%
      
      # change all words in string to title text
      str_to_title() %>%
      return()
  }

func_change_order <-
  function(df) {
    
    # please note this function is relatively simple, assuming a split between
    # character and numeric fields only.  the function can easily be amended, or
    # easier still is to coerce the calling dataframe fields to support the use
    # of this function
    df %>% 
      
      # select the calling dataframe character fields first
      select_if(is.character) %>% 
      cbind(
        
        # add numeric fields behind the character fields
        df %>% 
          select_if(is.numeric)
      ) %>%
      return()
  }

```

### Parameters and Configuration
The following code block sets configurable parameters, pointing to the source data file location and the decimal accuracy when parsing numbers.<br><br>

```{r}
var_path <- "../../resources/family_food_budget/"
options(digits=9)

```

### Import
I've previously created a spreadsheet with menu items, underlying ingredients and associated information.  Next we import each tab from the Excel file as per the `var_path` parameter into the `df_import` dataframe, nesting each tab in the `file` column.<br><br>

```{r}
df_import <-
  
  # list all files matching the pattern in the parametarised set location
  list.files(path = var_path,
             pattern = "^family.+",
             recursive = TRUE) %>%
  
  # move the list of files into a tibble, changing the default column name
  # `value` to file_name_ext, representing the file name and extention
  enframe(name=NULL) %>%
  rename(file_name_ext = value) %>%
  
  # complete the file path used by the read_excel function
  mutate(file_path = paste(var_path, file_name_ext, sep = "/")) %>%
  
  # read all file tabs for the file and unnest
  mutate_at("file_path", list(file_tab = map), readxl::excel_sheets) %>%
  unnest() %>%
  
  # import the file using the file_path and file_tab as parameters
  mutate(file = map2(file_path, file_tab, function(file_path, file_tab) {
    readxl::read_excel(path = file_path,
                       sheet = file_tab,
                       guess_max = 100000)
  })) %>%
  
  # remove redundant variables
  select(-file_path,-file_name_ext)

```

Let's inspect the dataframe with the imported data.<br><br>

```{r}
df_import
```

The dataframe `df_import` reveals two imported tabs, including `price_list` and `menu`.

The `price_list` data contains ingredients, units purchased according to the *packaging* and its purchase price.  

Tibble `menu` is a list of meals, listing the `Meal` time-of-day, including breakfast, lunch, dinner or a snack only. Each menu item (meal) contains a list of ingredients and units consumed in the associated menu meal.<br><br>

### Reference Period
Here we instantiate two reference tibbles, including the period or timeframe for which the meals are planned, and the meal time-of-day.  

Tibble `ref_days_join_meal` is a cartesian product of `ref_days`, expanding meal times for each day, as listed in the `ref_meal_time` tibble.<br><br>

```{r}
ref_days <-
  seq.Date(from = ymd("2019-06-01"),
           to = ymd("2019-06-30"),
           by = "day") %>% 
  enframe(name=NULL) %>% 
  rename(Day = value)

ref_meal_time <-
  c("Breakfast", "Snack", "Lunch", "Dinner") %>% 
  factor(ordered = TRUE) %>% 
  enframe(name=NULL) %>% 
  rename(Meal = value)

ref_days_join_meal <-
  ref_days %>%
  
  # expand each day with the 4 mealtimes, creating a dataframe with combinations
  # of all days and mealtimes.
  expand(Day, Meal = ref_meal_time$Meal) %>% 
  
  # extract some attributes from the days, including the weekday, and the first
  # day of the week `SOW`
  mutate(Weekday = wday(Day,  label = TRUE)) %>% 
  mutate(SOW = floor_date(Day, week_start = 1, unit = "week")) %>% 
  mutate_at("Weekday",
            ~fct_relevel(., "Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))

```

### Menu Items
In this section we combine the menu list and the price list for each menu's ingredients, creating a `Menu_Cost` column summing the total value of all underlying ingredients for each meal.  

Ingredients are nested within the `Ingredients` column for each meal.  The head the resulting nested table `df_menu_items` is printed below.<br><br>

```{r}
df_menu_items <-
  
  # unnest the menu nested file, containing each meal and its ingredients
  df_import %>%
  filter(file_tab == "menu") %>%
  unnest() %>%
  
  # coerse the meal to a factor
  mutate_at("Meal", factor, levels = ref_meal_time$Meal, ordered = TRUE) %>%
  
  # unnest and join the price list tibble by the Item, keeping the Item
  # (ingredient) and Price
  left_join(df_import %>%
              filter(file_tab == "price_list") %>%
              unnest() %>%
              select(Item, Price),
            by = "Item") %>%
  
  # calculate the cost of the ingredient, a product of the price and unit
  mutate(Cost = Price * Unit) %>%
  
  # calculate the total cost of the meal by summing the cost of all underlying
  # ingredients
  group_by(Meal, Menu) %>%
  mutate(Menu_Cost = sum(Cost)) %>%
  
  # regroup the the Meal, Menu adding Menu_Cost and nest ingredients
  group_by(Menu_Cost, add = TRUE) %>%
  nest(.key = "Ingredients") %>%
  
  # group by Meal and remove other groupings, adding a Meal_item row number
  group_by(Meal, add = FALSE) %>%
  mutate(Meal_item = row_number())

df_menu_items %>% 
  head() %>% 
  mutate_at("Menu_Cost", ~scales::dollar(., prefix = "£")) %>% 
  rename_all(func_present_headers) %>% 
  kable(format = "html") %>% 
  kableExtra::kable_styling(full_width = FALSE,
                font_size = 9)

```
<br><br>

### Function
The function `func_random_meal` randomises meals for each mealtime, the size matching the number of days by week (`SOW`) in the `ref_days` tibble.  

The function draws randomly from meals for each mealtime.  For example, dinners will only sample from meals classified as *dinner*.  

It avoids sampling `with replacement` from the available menu meals unless there are fewer meals than the calling days.<br><br>

```{r}
func_random_meal <-
  function(param_meal, param_days) {
    # create a parameter with the number of days in the param_days parameter
    param_sample_size <- count(param_days) %>% as.integer()
    
    # create a parameter, returning the count of items in the meal
    param_menu_item_size <- 
      df_menu_items %>%
      
      # filter by the calling param_meal
      filter(Meal == param_meal) %>%
      nrow()
    
    # function that returns TRUE if the param_menu_item_size <
    # param_sample_size, enabling the `tmp_meals` function to sample with
    # replacement
    param_replace <-
    if (param_menu_item_size < param_sample_size) {
      TRUE %>% return()
    } else {
      FALSE %>% return()
    }
    
    # Sample ----
    tmp_meals <-
      df_menu_items %>%
      
      # filter by the calling param_meal
      filter(Meal == param_meal) %>%
      
      # create a sample size according the the parameter param_sample_size,
      # enabling replacement if the param_menu_item_size < param_sample_size
      sample_n(size = param_sample_size, replace = param_replace) %>%
      ungroup() %>%
      
      # remove unecessary attributes
      select(-Meal,-Meal_item) %>%
      
      # add row number
      mutate(Day_order = row_number())

    # Join ----
    param_days %>%
      
      # join the newly created tibble `tmp_meals` to the param_days tibble,
      # using the Day_order as key to join
      mutate(Day_order = row_number()) %>%
      inner_join(tmp_meals, by = "Day_order") %>%
      return()
  }

```

### Randomise Meals
We start of by setting a seed, thus making results repeatable.  

Next, we create a new tibble `df_meal_plan` by passing the meals for each week to the `func_random_meal` function, generating and returning a random list of meals to the `Menu` column.

`Menu_Cost` summarises the total cost of all underlying ingredients for the meal.

Executing the code block outputs the head of the new tibble for inspection.<br><br>

```{r}
set.seed(567)
# set seed to make the results repeatable

df_meal_plan <-
  # for each day and mealtime combination
  ref_days_join_meal %>%
  
  # group by week and mealtime, and nest all days within it
  group_by(SOW, Meal) %>%
  nest(Weekday, Day) %>% 
  
  # map the meal and nested days to the `func_random_meal` function and return
  # result to the Menu column
  mutate(Menu  = map2(Meal, data, func_random_meal)) %>%
  
  # unnest the result and arrange the output
  unnest(Menu) %>%
  arrange(SOW, Weekday, Meal) %>% 
  mutate_at("SOW", ~ paste0(format(., "%yW%W: "),
                           format(., "%d %b - "), 
                           format(. + 6, "%d %b")) %>%
              factor(., ordered = TRUE))

# check the results by printing the first day worth of meals
df_meal_plan %>% 
  head(4) %>% 
  
  # change the date formatting for improved presentation
  mutate_if(is.Date, format, format = "%d-%b") %>% 
  
  # presentation options
  mutate_at("Menu_Cost", ~scales::dollar(., prefix = "£")) %>% 
  rename_all(func_present_headers) %>% 
  kable(format = "html") %>% 
  kableExtra::kable_styling(full_width = FALSE, font_size = 10)
```
<br><br>

### Month Lookahead
This section is really what this whole exercise is about!  

We've created random meals for a period, ready to print a copy of the *first* output and pin it to the family whiteboard.  We now have a menu plan for the month ahead - **result**.

The *second* output is used for budgeting the food bill once we have taken stock, and calculate what is to be purchased to satisfy the newly created budget plan.<br><br>

```{r}
# Meal ----
df_meal_plan %>% 
  select(SOW, Weekday, Meal, Menu) %>% 
  spread(Weekday, Menu, fill = "--") %>% 
  ungroup() %>% 
  mutate_at("SOW", format, format = "%yW%W") %>% 
  mutate_at("Meal", factor, levels = ref_meal_time$Meal) %>%
  arrange(SOW, Meal) %>%
  kable(format = "html", align = "c") %>% 
  kable_styling("striped", full_width = FALSE, font_size = 10) %>%
  column_spec(1, bold = TRUE) %>%
  collapse_rows(columns = 1:2, valign = "top")

# Money ----
df_meal_plan %>% 
  group_by(SOW, Weekday, Meal) %>% 
  summarise_at("Menu_Cost", sum, na.rm = TRUE) %>% 
  spread(Weekday, Menu_Cost) %>% 
  ungroup() %>% 
  mutate_at("SOW", format, format = "%yW%W") %>% 
  mutate_at("Meal", factor, levels = ref_meal_time$Meal) %>%
  arrange(SOW, Meal) %>% 
  func_create_summary() %>% 
  mutate_at(c("SOW", "Meal"), as.character) %>% 
  group_by_if(is.character) %>% 
  nest(.key = "Total") %>% 
  mutate(data = Total) %>% 
  mutate_at("Total", map, rowSums, na.rm = TRUE) %>% 
  unnest() %>%
  mutate_if(is.numeric, ~scales::dollar(., prefix = "£")) %>% 
  mutate_if(is.character, str_replace, pattern = "£NA", replacement = "--") %>% 
  kable(format = "html", align = "c") %>% 
  kable_styling("striped", full_width = FALSE, font_size = 10) %>%
  column_spec(1, bold = TRUE) %>%
  collapse_rows(columns = 1:2, valign = "top")
```

### Purchasing
The following section is creating *output* that helps to manage the newly created menu plan.

The newly created menu plans now enables one to project spend and even create a shopping list for each week[^1].<br><br>

```{r}
df_meal_plan_purchases <-
  df_meal_plan %>%
  unnest() %>%
  select(Day, Item, Unit, Price, Cost) %>%
  mutate(Meals = 1) %>%
  group_by(Item) %>%
  mutate_at(c("Unit", "Meals"), list(cum = cumsum)) %>%
  mutate_at("Unit_cum", ceiling) %>%
  mutate_at("Unit_cum", list(buy = ~. != lead(.),
                             buy_volume = ~lead(.) - .)) %>%
  mutate(Cost_cum = Unit_cum * Price) %>%
  mutate_at("Day",
            list(SOW = floor_date),
            unit = "week",
            week_start = 1) %>% 
  group_by(SOW, add = TRUE) %>% 
  summarise_at("buy_volume", sum, na.rm = TRUE)

df_plot <-
  df_meal_plan_purchases %>% 
  inner_join(
    df_import %>%
      filter(file_tab == "price_list") %>%
      unnest(),
    by = "Item"
  ) %>% 
  filter(buy_volume > 0) %>% 
  mutate(Cost = Price * buy_volume) %>% 
  ungroup() %>% 
  mutate(Total_Cost = sum(Cost, na.rm = TRUE)) %>% 
  arrange(desc(Total_Cost)) %>% 
  mutate_at("Item", fct_inorder) %>% 
  group_by(SOW, add = FALSE) %>% 
  mutate_at("Cost", list(Cost_Cum = cumsum)) %>% 
  ungroup() %>% 
  arrange(SOW) %>% 
  mutate_at("SOW", ~ paste0(format(., "%yW%W: "),
                           format(., "%d %b - "), 
                           format(. + 6, "%d %b")) %>%
              factor(., ordered = TRUE))

df_plot %>% 
  ggplot(aes(x = Item, y = Cost, fill = Cost)) +
  geom_col() +
  geom_line(aes(y = Cost_Cum, group = SOW)) +
  ggrepel::geom_label_repel(
    data = df_plot %>%
      group_by(SOW, add = FALSE) %>%
      top_n(n = 1, wt = Cost_Cum),
    aes(y = Cost_Cum, label = scales::dollar(Cost_Cum, prefix = "£")),
    fill = "white"
  ) +
  facet_grid(SOW ~ ., switch = "y") +
  scale_y_continuous(labels = scales::dollar_format(prefix = "£")) +
  scale_fill_continuous(labels = scales::dollar_format(prefix = "£")) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    strip.placement = "outside",
    strip.text.y = element_text(angle = 180, vjust = 1),
    strip.background.y = element_blank(), 
    legend.position = "bottom",
    legend.key.width=unit(2.5,"cm")
  ) +
  labs(title = "Montly shopping costs by week",
       y = "Purchases")
```
<br><br>

### The Shopping List
Being able to estimate my food bill is a result.  However, having a prospective shopping list is practical and useful.<br><br>

```{r , echo=TRUE, message=FALSE, warning=FALSE, results='asis', paged.print=TRUE}

func_output_group_item_table <-
  function(param_group, param_group_selector) {
    # print the distinct grouping category as header
    cat(paste0(
      "<h4>",
      str_replace_all(param_group, pattern = "_", replacement = " "),
      "</h4>"
    ))
    
    df_plot <-
      df_plot %>%
      select(SOW, Item, volume = buy_volume, Price, Cost, Cost_Cum)
    
    df_plot %>%
      # filter the master data by the grouping category parameter
      filter(!!param_group_selector == param_group) %>%
      ungroup() %>%
      
      #  remove the grouping category from it
      select(-!!param_group_selector) %>%
      
      # create a row with a sum total for each numeric column
      mutate_at("Cost_Cum", as.character) %>% 
      func_create_summary() %>%
      
      # styling and presentation options executed next, as described in the
      # previous code block
      mutate_at("volume", as.character) %>% 
      mutate_at("Cost_Cum", as.numeric) %>% 
      rename_all(func_present_headers) %>%
      mutate_if(is.numeric, ~ round(., 2) %>% scales::dollar(., prefix = "£")) %>%
      mutate_if(is.character, str_remove, pattern = "\\£NA") %>%
      mutate_if(is.factor,
                str_replace_all,
                pattern = "_",
                replacement = " ") %>%
      kable(align = "r", format = "html") %>%
      kableExtra::kable_styling(full_width = FALSE,
                                font_size = 12) %>%
      kableExtra::collapse_rows(columns = 1:4, valign = "middle") %>%
      return()
  }

func_output_group_item_table_iterator <-
  function(param_group_selector) {
    for (param_group in (df_plot %>%
                         distinct(!!param_group_selector) %>%
                         pull()))  {
      func_output_group_item_table(param_group, param_group_selector) %>% print()
    }
  }

func_output_group_item_table_iterator(quo(SOW))
```

### EDA
This section explores our menu by mealtime, visualising the ranked cost of menus.<br><br>

```{r}
df_meal_plan %>% 
  distinct(Meal, Menu, Menu_Cost) %>% 
  arrange(Meal, desc(Menu_Cost)) %>% 
  mutate_at("Menu", fct_inorder) %>% 
  ggplot(aes(x = Menu, y = Menu_Cost, fill = Menu_Cost)) +
  geom_col() +
  geom_label(aes(label = format(Menu_Cost, digits = 2, big.mark = ",", format = "#.##") %>%
                   paste0("£", .)),
             col = "white", 
             size = 3) +
  coord_flip() +
  theme_classic() +
  scale_y_continuous(labels = scales::dollar_format(prefix = "£")) +
  facet_grid(Meal ~ ., 
             scales = "free_y", 
             space = "free", 
             switch = "y") +
  theme(strip.placement = "outside",
        strip.text.y = element_text(angle = 180, 
                                    face = "bold", 
                                    size = 10,
                                    hjust = 0, 
                                    vjust = 1), 
        strip.background.y = element_blank(),
        plot.subtitle = element_text(size = 8, face = "italic"),
        legend.key.height=unit(0.5,"line"),
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 8),
        legend.position = "bottom") +
  labs(title = "Cost of meals", 
       subtitle = "Grouped by mealtime - Ordered by Cost",
       fill = "Cost/ Meal",
       y = "Cost/ Meal",
       x = "")

```

The following visualisation contrasts a few meals to provide a side-by-side breakdown of costs per menu.  It is clear that using meat and complexity is what increases the cost of meals.<br><br>

```{r}
# Expensive Items ----
df_menu_items %>%
  filter(grepl("curry|roast|spag|carb", Menu, ignore.case = TRUE)) %>%
  unnest() %>%
  arrange(desc(Cost)) %>%
  mutate_at("Item", fct_inorder) %>%
  ggplot(aes(Menu, Cost, fill = Item)) +
  geom_col(position = "stack", show.legend = FALSE) +
  geom_label(
    aes(
      label = format(
        Cost,
        digits = 2,
        big.mark = ",",
        format = "#.##"
      ) %>% paste0("£", .) %>% paste(Item, .)
    ),
    position = "stack",
    col = "white",
    size = 3,
    show.legend = FALSE
  ) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "£")) +
  theme_classic() +
  labs(title = "Comparing the cost of ingredients for selected meals")
```

## Summary
This post doesn’t describe life-changing methods or insights. However, it demonstrates that coding can be used to solve simple, everyday problems, even having a meaningful impact on my little family.  **Imagine though what it can do for seemingly insignificant things in business.**

<hr>

## Session

```{r}
sessionInfo()

```


[^1]: <sub>assuming that the cupboard is one unit up on all required items</sub>

<!-- <div class="fab"> -->
<!-- <script type="text/javascript"> -->

<!-- // toggle visibility of R source blocks in R Markdown output -->
<!-- function toggle_R() { -->
<!--   var x = document.getElementsByClassName('r'); -->
<!--   if (x.length == 0) return; -->
<!--   function toggle_vis(o) { -->
<!--     var d = o.style.display; -->
<!--     o.style.display = (d == 'block' || d == '') ? 'none':'block'; -->
<!--   } -->

<!--   for (i = 0; i < x.length; i++) { -->
<!--     var y = x[i]; -->
<!--     if (y.tagName.toLowerCase() === 'pre') toggle_vis(y); -->
<!--   } -->

<!--     var elem = document.getElementById("fab"); -->
<!--     if (elem.value === "Hide Code Blocks") elem.value = "Show Code Blocks"; -->
<!--     else elem.value = "Hide Code Blocks"; -->
<!-- } -->


<!-- document.write('<input onclick="toggle_R()" type="button" id="fab" value="Hide Code Blocks"</input>') -->

<!-- </script> -->
<!-- </div> -->

