---
title: dplyr 0.8.1 grouping functions update
author: Gerhard Groenewald
date: '2019-06-06'
slug: dplyr-0-8-1-update-to-grouping-functions
categories: [R]
tags: [R, tips tricks, library, review]
subtitle: 'Changes to group_modify() and group_map()'
---

[RStudio](https://www.rstudio.com) has just released a [minor update](https://www.tidyverse.org/articles/2019/05/dplyr-0-8-1/) to [dplyr](https://dplyr.tidyverse.org).  They had a rethink of new [grouping *purrr-style* functions](https://dplyr.tidyverse.org/reference/group_map.html) used to iterate on grouped tibbles.

The changes include:

- `group_map()` is now used for iterating on grouped tibbles.  It however makes no assumptions about the return type of each operation, combining results in a list - similar to `purrr::map()`.
- The previous behaviour was renamed to `group_modify()`, always returning and combining grouped tibbles by evaluating each operation with a reconstructed grouping structure - similar to `purrr::modify()`.<br><br>

An extract from the wiki states that:

> group_modify() is an evolution of do(), if you have used that before.

A typical implementation changes from `do(function(.$column))` to `group_modify( ~ {function(.$column)})`, both yielding the same result.  `group_map` similarly applies a function across each grouped tibble, albeit dropping the grouping variables and outputting lists for each group.

Let's prepare data for the brief demonstration.<br><br>

```{r message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)
library(tidyr)
library(magrittr)
library(conflicted)

# Implementing conflicted
purrr::walk(c("filter", "lag", "intersect", "setdiff", "setequal", "union"), function(func) {
  conflict_prefer(func, "dplyr")
})
```

```{r, include=FALSE}
func_tidy_present <-
  function(data, header = NA, return_row_count = 4) {
    
    cat(paste0("<br><h5>", header, "</h5>"))
    
    data %>% 
      ungroup() %>%
      head(return_row_count) %>% 
      mutate_if(is.numeric, round, 2) %>% 
      knitr::kable(format = "html") %>%
      kableExtra::kable_styling(full_width = FALSE, position = "left", font_size = 12) %>% 
      return()
  }
```

```{r, echo=TRUE, message=FALSE, warning=FALSE, results='asis', paged.print=TRUE}
iris_grouped <-
  iris %>% 
  
  # Make the table long and thin by collapsing numeric attributes and associated
  # values describing Sepal and Petal lengh and width into the `Field` and `val`
  # columns respectively.
  gather(Field, val, -Species) %>% 
  
  # group by Species and Field to execute an operation over the val data for
  # each respective group
  group_by(Species, Field)

iris_grouped %>%
  
  # The function returns the first 4 rows of the dataframe, formatting the
  # output to HTML using knitr and kableextra, printing a passed header as shown
  # below
  func_tidy_present(., 
                    header = "Tidied *Gathered & Grouped* Dataframe")
```

<br><hr><br>

The demonstration below shows the subtle change, introducing more syntactic cohesion to the coding flow.<br><br>

```{r, echo=TRUE, message=FALSE, warning=FALSE, results='asis', paged.print=TRUE}
iris_grouped %>%
  
  # Apply the function for group using the do method
  do(psych::describe(.$val)) %>%
  
  # Tidyup to present
  func_tidy_present(., header = "Result of the *Do* method")

iris_grouped %>% 
  
  # Apply the function for group using the Group_Modify method
  group_modify( ~ {psych::describe(.$val)}) %>% 
  
  # Tidyup to present
  func_tidy_present(., header = "Result of the *Group Modify* method")
```

<br><hr><br>

An implentation of the `quantile` function returns a list that is subsequently `enframed` and spread before presentation.<br><br>

```{r, echo=TRUE, message=FALSE, warning=FALSE, results='asis', paged.print=TRUE}
iris_grouped %>%
  
  # Apply the function for group using the Group_Modify method
  group_modify( ~ {
    quantile(.x$val, probs = c(0.25, 0.5, 0.75)) %>%
      tibble::enframe(name = "prob", value = "quantile")
  }) %>%
  spread(Field, quantile) %>%
  
  # Tidyup to present
  func_tidy_present(.,
                    header = "Result of the *Group Modify* method applying the *quantile* function",
                    return_row_count = 9)
```